Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-29 paper/main.tex
rewrite imaging-based metric subsection

\subsubsection{Imaging-based Comparison Algorithm}
\label{subsec:imaging-based-comparison}
The preceding algorithm is performant and effective but can not be applied in the case when a
division event splits apart the bacterium since this would involve the comparison of two sets of
vertices with only one.
This configuration is displayed in figure~\ref{fig:mask-difference-metric} (E) where two distinct
bacterial configurations originate from the same initial state but contain different amounts of
bacteria.
This mismatch can happen, e.g. due to non-matching division times.
Thus a cell which is present in the simulation might not yet have a corresponding cell in the
experimental data at this time point yet (or vice-versa).
To explain why the direct skeleton-space metric can not be reused on confined
intervals which do not include division events and for further details, we refer the interested
reader to the supplementary material~\ref{sec:supplement-comparison-algorithms}.

In order to be able to compare the differing cellular configurations, we require additional
methods which can be used in time intervals containing cellular divison events.
To this end we use the numerical results of the \ac{abm} simulation and our visualization
algorithms to generate cell-masks (see section~\ref{subsection:visualization}).
Since each pixel has a unique color value which is directly related to one particular cell or
background (black), we compare the generated masks to the experimentally obtained ones pixel by
pixel.
Whenever two colors are not matching, a penalty is assigned for each pixel.
The total cost is the sum of all penalties.
% This algorithm is an implementation of a signed area measure.
It is important to take into account the parental relationship between the cells, by checking if
either one of the colors corresponding to the compared colors is a daughter of the other cell.
Figure~\ref{fig:mask-difference-metric} (A,B) shows two successive states of a numerical simulation
with varying amounts of cells.
We can see that some cells have divided while others have only grown without completing a division
step.
The resulting penalties without accounting for the parental relationships are displayed in
Figure~\ref{fig:mask-difference-metric} (C).
Black indicates that the pixels are matching and no penalty is assigned while a white pixel
indicates that the original pixels are not matching, thus adding to the overall cost.
In contrast, Figure~\ref{fig:mask-difference-metric} (D) takes the parental relationship into
account and highlights regions where compared pixels belong to mother-daughter cells in gray.
We assign a penalty $0\leq p \leq 1$ which is a hyperparameter for this optimization algorithm.
% It is possible to assign a custom value for the parental cost value.
Settings its value to $0$ results in a contiuous cost function in time.

The developed optimization scheme was mainly driven by technical challenges, but it is also
biologically motivated.
In the short time period after the division event, the two resulting daughter cells will cover
the space that the mother had taken up.
Often times, the exact time at which the fission event has commenced can not be determined
accurately, thus leading to uncertainties with respect to the question when the life of the daughter
cells begins and the one of the mother ends.
These uncertainties which are also common problems for segmentation methods, further highlight the
importance of this comparison scheme.

\subsubsection{Testing the Comparison Algorithms with Division Events}
When utilizing a computational model (see section~\ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the cost between two successive iteration steps
$t_n$ and $t_{n+1}$.
We expect to obtain an exponential growth in the cost.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area with
respect to time, which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (F) shows the total cost for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.

@@

\section{Author Contributions}
\textbf{Conceptualization:} Jonas Pleyer\\
\textbf{Investigation:} Jonas Pleyer\\
\textbf{Methodology} Jonas Pleyer\\
\textbf{Software:} Jonas Pleyer\\
\textbf{Supervision:} Christian Fleck\\
\textbf{Writing – original draft:} Jonas Pleyer\\
\textbf{Writing – review \& editing:} Christian Fleck \& Jonas Pleyer

%###################################################################################################
\section{Data and Code Availability}
\label{section:data-code-availability}
All data and code which has been used in this study is available at the github repository
\url{https://github.com/jonaspleyer/cr\_mech\_coli}.
It includes detailed descriptions of how to obtain the used data and bring it into a usable format.
The package which bundles the developed funtionality is published at
\url{https://pypi.org/project/cr-mech-coli/} from which it is readily available as a python
dependency.\\
The time series of microscopic images which has been studied in section~\ref{section:application-2}
was extracted from individual frames of a video~\cite{Trude1982} that
showcases the exponential growth of bacteria.

>>>>

\subsubsection{Imaging-Space Metric}
\label{subsec:image-space-metric}
\paragraph{Motivation}
The preceding algorithm is performant and effective but can not be applied in the case when a
division event splits apart the bacterium since this would require comparing sets of unequal
cardinality.
This situation is displayed in figure~\ref{fig:mask-difference-metric} (E) where two distinct
bacterial configurations originate from the same initial state but contain varying numbers of
bacteria at the next time step.
Such a mismatch can happen, e.g. due to non-matching division times of model and data.
Thus a cell which is present in the simulation might not yet have a corresponding cell in the
experimental data at this time point yet (or vice-versa).

In addition to these technical challenges come practical ones.
Within a short time period after the division event, the two resulting daughter cells will cover
the space that the mother had taken up.
However, the exact time point for the fission event can not be determined accurately in practice,
thus leading to natural uncertainties with respect to the question when the lives of the daughter
cells begin and the one of the mother ends.
Such uncertainties which are also common problems for segmentation methods, further highlight the
importance of a purely imaging-based comparison scheme.

These thoughts explain why the direct metric can not be simply reused on confined intervals which
do not include division events:
It does not account for the inaccuracy of the division timing.
Furthermore, finding such confined intervals becomes increasingly difficult for larger numbers of
bacteria where division events occur much more frequently.
This fundamental limitation will be circumvented by the imaging-based method which we propose.

\paragraph{The Metric}
We develop an approach which can be understood as a metric in image space where we quantify
differences in the visual footprint of two bacterial configurations.
Within the realm of image segmentation many metrics such as the Jaccard Index~\cite{Murphy1996},
Dice-Sørensen Coefficient~\cite{Dice1945}, Precision and Recall~\cite{Takaya2015} or F1
Score~\cite{Taha2015} are actively used to deal with similar questions.
However, these statistical methods are all based upon the notion of identifying matching and
non-matching regions.
It is therefore first necessary to define a criterion for this.

% To compare the differing cellular configurations, we require additional
% methods which can be used in time intervals containing cellular divison events.
We can use the previously introduced \nameref{subsection:visualization} methods to obtain cell-masks
from the numerical results of the \ac{abm} simulation.
Since we utilize an in-image encoding scheme, each pixel has a unique color value which is directly
related to one particular cell or background (black).
We can now compare the two masks pixel by pixel.
Whenever two colors are not matching, a penalty is assigned for each point in the pixel-grid.
The total cost is the sum of all penalties.
With this approach it is important to take into account the parental relationship between the cells,
by checking if either one of the colors corresponding to the compared cells is a daughter of the
other cell.
Figure~\ref{fig:mask-difference-metric} (A,B) shows two states of a numerical simulation
with varying amounts of cells.
We can see that some cells have divided while others have only extended in the same time interval.
The resulting penalties without accounting for the parental relationships are displayed in
Figure~\ref{fig:mask-difference-metric} (C).
Black indicates that the pixels are matching and no penalty is assigned while white
indicates that the original pixels are not matching, thus adding to the overall cost.
In contrast, Figure~\ref{fig:mask-difference-metric} (D) takes the parental relationship into
account and highlights regions where compared pixels belong to mother-daughter cells in gray.
We assign a specific parent penalty $0\leq p\leq 1$ value for these points which should be seen as
an additional hyperparameter for this optimization algorithm.
% It is possible to assign a custom value for the parental cost value.
Settings its value to $0$ results in a contiuous cost function in time.
In the future, more complex choices for this parent penalty might be feasible such as
time-dependence $p(t-t_\text{division})$.

While our primary concern is to compare numerically obtained results with experimental data,
this approach allows us to compare all spatial configurations which can be represented
by segmented microscopic images irrespective of how these are obtained.
On a higher level, this means that our method is agnostic to the underlying cellular representation.
This means that other models who even utilize very distinct cellular representations but are able to
render segmented images can reuse our method.

\subsubsection{Testing the Comparison Algorithms with Division Events}
When utilizing a computational model (see \ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the cost between two successive iteration steps
$t_n$ and $t_{n+1}$.
We expect to obtain an exponential growth in the cost.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area with
respect to time, which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (F) shows the total cost for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.

@@

%###################################################################################################
\section{Acknowledgements}
\subsection{Author Contributions}
\textbf{Conceptualization:} Jonas Pleyer\\
\textbf{Investigation:} Jonas Pleyer\\
\textbf{Methodology} Jonas Pleyer\\
\textbf{Software:} Jonas Pleyer\\
\textbf{Supervision:} Christian Fleck\\
\textbf{Writing – original draft:} Jonas Pleyer\\
\textbf{Writing – review \& editing:} Christian Fleck \& Jonas Pleyer

\subsection{Declaration of AI Use}
The authors acknowledge the use of Gemini (Google) for assistance with language editing and
proofreading.
In particular, no generative aspect was used and each suggested change was carefully reviewed and
utilized only as additional feedback source.
All contents of the paper are written and audited by the authors which is why they take full
responsibility for the contents presented in this publication.

\section{Data and Code Availability}
\label{section:data-code-availability}
All data and code which has been used in this study is available at the github repository
\url{https://github.com/jonaspleyer/cr\_mech\_coli}.
It includes detailed descriptions of how to obtain the used data and bring it into a usable format.
The package which bundles the developed funtionality is published at
\url{https://pypi.org/project/cr-mech-coli/} from which it is readily available as a python
dependency.\\
The time series of microscopic images which has been studied in section~\ref{section:application-2}
was extracted from individual frames of a video~\cite{Trude1982} that
showcases the exponential growth of bacteria.
